---
title: "Gapminder Dataset Analysis"
date: "Dec 13, 2022"
output: 
  html_document:
    theme: paper
    toc: true
    toc_float: true
---

<style type="text/css">
  h1, h2, h3, h4 {
    color: #2196F3;
    text-align: center;
  }
  body{
  font-size: 12pt;
  }
</style>

<center>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE,
  warning = FALSE
)
options(scipen = 999) # to prevent scientific notation on graph axes
```

```{r}
library(tidyverse)
library(plotly)
library(ggpubr)
library(rstatix)
library(DT)
```

```{r}
gapminder <- read_csv("gapminder_clean.csv",
  col_select = c(2:20),
  col_names = c(
    "row", "country", "year", "agriculture_value_added",
    "co2_emissions", "domestic_credit_by_financial_sector",
    "electric_power_consumption", "energy_use", "exports",
    "fertility_rate", "GDP_growth", "imports",
    "industry_value_added", "inflation", "life_expectancy",
    "population_density", "services_value_added", "population",
    "continent", "GDP_per_cap"
  ), skip = 1
)
```

### CO2 Emissions Versus GDP Per Capita In 1962

```{r}
co2_GDP_1962 <- gapminder %>%
  filter(year == 1962) %>%
  select(co2_emissions, GDP_per_cap) %>%
  na.omit()

# check normality
fig1 <- plot_ly(
  x = co2_GDP_1962$co2_emissions,
  type = "histogram"
) %>%
  layout(
    xaxis = list(title = "CO2 Emissions in 1962"),
    yaxis = list(title = "Count")
  )
fig2 <- plot_ly(
  x = co2_GDP_1962$GDP_per_cap,
  type = "histogram"
) %>%
  layout(
    xaxis = list(title = "GDP Per Capita in 1962"),
    yaxis = list(title = "Count")
  )
fig <- subplot(fig1, fig2, titleX = TRUE, titleY = TRUE) %>%
  layout(title = "Histograms For Checking Normality", showlegend = FALSE)
fig
```

```{r}
corr_1962 <- cor.test(co2_GDP_1962$co2_emissions,
  co2_GDP_1962$GDP_per_cap,
  method = "spearman"
)
rho_1962 <- round(corr_1962$estimate, 2)
p_val_1962 <- corr_1962$p.value
```

Data was not normally distributed so Spearman's correlation was used to 
look at the relationship between CO2 emissions and GDP per capita in 1962. The 
correlation coefficient was `r rho_1962` with a p-value of `r p_val_1962`.

```{r}
ggplot(
  co2_GDP_1962,
  aes(x = GDP_per_cap, y = co2_emissions)
) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "CO2 Emissions Versus GDP Per Capita in 1962",
    y = "CO2 emissions \n (metric tons per capita)", x = "GDP per capita"
  ) +
  annotate("text", x = 1000, y = 1000, label = paste("R = ", rho_1962)) +
  annotate("text", x = 1000, y = 500, label = paste("p-value = ", p_val_1962))
```

### CO2 Emissions Versus GDP Per Capita For All Years
```{r}
# check normality
fig1 <- plot_ly(
  x = gapminder$co2_emissions,
  type = "histogram"
) %>%
  layout(
    xaxis = list(title = "CO2 Emissions"),
    yaxis = list(title = "Count")
  )
fig2 <- plot_ly(
  x = gapminder$GDP_per_cap,
  type = "histogram"
) %>%
  layout(
    xaxis = list(title = "GDP Per Capita"),
    yaxis = list(title = "Count")
  )
fig <- subplot(fig1, fig2, titleX = TRUE, titleY = TRUE) %>%
  layout(title = "Histograms For Checking Normality", showlegend = FALSE)
fig
```

```{r}
co2_gdp_corr <- gapminder %>%
  select(year, co2_emissions, GDP_per_cap, year) %>%
  na.omit() %>%
  group_by(year) %>%
  summarise(
    correlation = cor(co2_emissions, GDP_per_cap, method = "spearman")
  )

viz <- ggplot(co2_gdp_corr, aes(x = year, y = correlation)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(
    breaks = seq(1960, 2010, by = 5),
    minor_breaks = seq(1960, 2010, by = 1)
  ) +
  labs(
    title = "Correlation Between CO2 Emissions and GDP",
    y = "Spearman Correlation \n Between CO2 Emissions and GDP", x = "Year"
  )
ggplotly(viz)
```

```{r}
max_corr_year <- co2_gdp_corr %>%
  filter(correlation == max(correlation)) %>%
  select(year)
```

Data was not normally distributed so a Spearman's correlation was used. 
Correlation between CO2 emissions and GDP per capita was highest in 
`r max_corr_year`.

### CO2 Emissions Versus GDP Per Capita In `r max_corr_year`

```{r}
data_from_max_year <- gapminder %>%
  filter(year == max_corr_year$year) %>%
  select(co2_emissions, GDP_per_cap, population, continent) %>%
  na.omit()
```

```{r}
viz <- ggplot(
  data_from_max_year,
  aes(
    x = GDP_per_cap, y = co2_emissions,
    size = population, color = continent
  )
) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = paste("CO2 Emissions vs GDP Per Capita in ", max_corr_year),
    y = "CO2 emissions (metric tons per capita)", x = "GDP per capita"
  )
ggplotly(viz)
```

### Energy Use By Continent

Before comparing energy use in different continents, a quantile-quantile 
plot was used to check the normality of energy use data.

```{r}
continent_energy_df <- gapminder %>%
  select(energy_use, continent) %>%
  na.omit() %>%
  mutate(continent = as.factor(continent))

# test for normality
ggqqplot(continent_energy_df, "energy_use",
  xlab = "Theoretical Quantiles",
  ylab = "Sample Quantiles", main = "Normal Q-Q Plot"
)
```

Because the data was not normally distributed, a Kruskal-Wallis test
and Wilcoxon test was used was used to compare the average energy 
use in different continents since 1962. 

```{r}
continent_energy_post_hoc <- pairwise_wilcox_test(continent_energy_df,
  energy_use ~ continent,
  paired = FALSE
) %>%
  filter(p.adj < 0.05) %>%
  add_xy_position(x = "continent")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
continent_energy_df %>%
  ggplot(aes(x = continent, y = energy_use)) +
  geom_boxplot() +
  stat_compare_means(
    method = "kruskal.test",
    label.x.npc = "left", label.y.npc = "top"
  ) +
  labs(
    title = "Energy Use by Continent",
    y = "Energy Use (kg of oil equivalent per capita)", x = "Continent"
  ) +
  stat_pvalue_manual(continent_energy_post_hoc)
```

There were significant differences in energy use between continents, with 
Oceania having the highest consumption per capita.

### Imports In Europe & Asia After 1990

A quantile-quantile plot was used to check the normality of imports data
in Europe and Asia after 1990. 

```{r}
imports_eu_asia <- gapminder %>%
  filter(continent == "Europe" | continent == "Asia") %>%
  filter(year > 1990) %>%
  select(continent, year, imports) %>%
  mutate(continent = as.factor(continent)) %>%
  na.omit()

# check if distribution is normal
ggqqplot(imports_eu_asia, "imports",
  xlab = "Theoretical Quantiles",
  ylab = "Sample Quantiles", main = "Normality Q-Q Plot"
)
imports_normality <- shapiro.test(imports_eu_asia$imports)
import_normality_p_val <- formatC(imports_normality$p.value,
  format = "e", digits = 2
)
```

A Shapiro-Wilk test was also used. The p-value was `r import_normality_p_val` 
which is less than 0.05 which means that the data is not normally distributed.
Therefore, a Wilcoxon rank-sum test was used compare imports between Europe 
and Asia.

```{r results=FALSE}
imports_eu_asia %>%
  ggplot(aes(x = continent, y = imports)) +
  geom_boxplot() +
  facet_wrap(~year) +
  stat_compare_means(
    method = "wilcox.test", label = "p.signif",
    comparisons = list(c("Europe", "Asia"))
  ) +
  labs(
    x = "Continent", y = "Imports of Goods and Services (% of GDP)",
    title = "Imports of Goods and Services in Europe and Asia"
  )
```

There was no significant difference between Europe and Asia (p>0.05) 
with respect to the percentage of GDP that accounts for imports of
goods and services in any years since 1990.

### Countries With High Population Density

This table shows the top 5 countries with the highest average ranking in 
population density across all time points.

```{r}
# top 5 countries with the highest population density rank
top_5_pop_density_countries <- gapminder %>%
  select(country, population_density, year) %>%
  na.omit() %>%
  group_by(year) %>%
  mutate(pop_density_rank = rank(population_density)) %>%
  ungroup() %>%
  group_by(country) %>%
  summarise(avg_pop_density_rank = mean(pop_density_rank, na.rm = TRUE)) %>%
  arrange(desc(avg_pop_density_rank)) %>%
  head(5)

datatable(top_5_pop_density_countries,
  colnames = c("Country", "Average Ranking of Population Density"),
  options = list(dom = "t"), # only shows table, not filters
  width = "60%"
)
```

The two countries with the highest average ranking in population density were
`r top_5_pop_density_countries$country[1]` and `r top_5_pop_density_countries$country[2]`.

### Countries With The Greatest Increase In Life Expectancy

The table below shows countries with the greatest incresease in life expectancy
since 1962.

```{r}
# Some countries experienced a decrease in life expectancy and then an increase.
# So, instead of simply subtracting the min from the max, I subtracted
# life expectancy in 1962 from life expectancy in 2007.
top_life_exp_diff <- gapminder %>%
  filter(year == 1962) %>%
  select(country, life_expectancy) %>%
  rename(life_exp_1962 = life_expectancy) %>%
  left_join(
    gapminder %>%
      filter(year == 2007) %>%
      select(country, life_expectancy) %>%
      rename(life_exp_2007 = life_expectancy),
    by = "country"
  ) %>%
  mutate(life_expectancy_diff = round(life_exp_2007 - life_exp_1962, 2)) %>%
  select(country, life_expectancy_diff) %>%
  arrange(desc(life_expectancy_diff)) %>%
  head(5)
```

```{r}
datatable(top_life_exp_diff,
  colnames = c("Country", "Increase In Life Expectancy (Years)"),
  options = list(dom = "t"), # only shows table, not filters
  width = "60%"
)
```

The country with the greatest increase in life expectancy since 1962
was `r top_life_exp_diff$country[1]` with an increase of 
`r top_life_exp_diff$life_expectancy_diff[1]` years.

</center>
